---
title: Calculating Returns for Two Kinds of Servicing Fee Schemes (Supplement)
author: Allen Sirolly
date: '2017-09-08'
slug: calculating-returns-for-two-kinds-of-servicing-fee-schemes-supplement
categories: []
tags: []
---

_This post is a supplement to [A Note on Lending Club and Institutional Investment](https://sirallen.name/blog/note-on-lending-club/)._

Suppose we'd like to compute ROI for a loan investment based on two different fee schemes. The first (call it $R_1$) charges a fee $s_1$ in proportion to each period's outstanding principal balance, while the second ($R_2$) charges a fee $s_2$ on all payments by the borrower. The borrower can "prepay" a loan with term $T$ by paying off the entire principal in $m < T$ months. Given initial principal $P_0$, monthly interest rate $r$, payment periods $m$, and payment sequence $\{A_i\}_{i=1}^m$, the principal amortizes according to
$$
P_i = (1+r)P_{i-1} - A_i = (1+r)^i P_0 - \sum_{j=1}^i (1+r)^{j-1} A_j
$$
and the corresponding annualized returns are
$$
R_1 = \left[ 1 + \frac{\sum_{i=1}^m (A_i - s_1 P_i) - P_0}{P_0} \right]^{12/m} - 1 \\
R_2 = \left[ 1 + \frac{\sum_{i=1}^m (1 - s_2) A_i - P_0}{P_0} \right]^{12/m} - 1.
$$
First assuming constant monthly payments given $m$,
$$
A_i = P_0 \frac{r(1+r)^m}{(1+r)^m - 1} \equiv A,
$$
we can use the following code to see how $R_1$ and $R_2$ vary with $m$. Note that by construction, i.e., by the choice of $\{A_i\}$, we have $P_m = 0$. Also note that $P_0$ isn't needed below, though I've kept it to make the calculations more explicit.

```{r}
A = function(P0, r, m) {
  # P0 -- initial principal
  # r  -- interest rate (monthly)
  # m  -- number of payment periods
  P0 * r * (1 + r)^m / ((1 + r)^m - 1)
}

P_i = Vectorize(
  # Compute outstanding principal after i payments
  function(A, P0, r, i) {
    if (i==0) return(P0)
    (1 + r)^i * P0 - A * sum((1 + r)^(0:(i-1)))
  },
  vectorize.args = 'i')

R1 = Vectorize(
  function(s, P0, r, m) {
    # s -- service fee rate (monthly, on remaining principal)
    # P0, r, m -- args to A()
    A_ = A(P0, r, m)
    ROI = (sum(A_ - s * P_i(A_, P0, r, 1:m)) - P0) / P0
    # annualize
    (1 + ROI)^(12/m) - 1
  },
  vectorize.args = 'm')

R2 = Vectorize(
  function(s, P0, r, m) {
    # s  -- service fee rate (monthly, on payments)
    # P0, r, m -- args to A()
    ROI = (m * (1 - s) * A(P0, r, m) - P0) / P0
    # annualize
    (1 + ROI)^(12/m) - 1
  },
  vectorize.args = 'm')
```

The test case below is a loan with \$1000 principal (again, doesn't matter) and 15% annual interest rate, with fees $s_1$ and $s_2$ equal to 1.3% per annum and 1%, respectively. (According to Lending Club's policy for whole loans, $s_2$ is variable but 1.3% per annum is the _highest_ servicing fee it will charge.)
```{r}
# Returns for institutional (whole loan) investors
r_inst = R1(s=.013/12, P0=1000, r=.15/12, m=1:36)

# Returns for retail (Note) investors
r_ret  = R2(s=.01, P0=1000, r=.15/12, m=1:36)

matplot(1:36, cbind(r_inst, r_ret), las=1, type='l',
        xlab='Months until full payment (m)',
        ylab='Annualized returns')
legend('topright', c('R1','R2'), lty=c(1,2), col=c('black','red'),
       inset=.01)
```

This isn't a complete picture since it doesn't account for returns from reinvestment (purchasing new Notes or whole loans after receiving each $A_i$, which can be done recursively).^[For `$m = 1$` annualization and reinvestment in "identical" loans are effectively the same. But for `$m > 1$`, the return calculation doesn't account for the additional returns that can be achieved through reinvestment.] But it's clear that investors who are charged fees under scheme $R_1$ can achieve higher returns with early payment, i.e., `$ \Delta_m R_1 < 0$`. In contrast, early payment is a suboptimal outcome under scheme $R_2$.

We can't yet declare the matter settled as there's a potential problem with the above graph. Empirically, given $m$, the borrower tends to back-load prepayment to later periods, so the assumption of fixed payments isn't very realistic. (I checked this using Lending Club's payments data, but I'll defer the evidence to a future post.) A more realistic flow of payments would be constant installments for $m-1$ periods (computed with the formula for $A_i$ using the loan term $T$ in place of $m$, which gives the _minimum_ amount the borrower must pay each month), and one large payment to expunge all outstanding principal (with interest) in period $m$.

It's straightforward to modify some of the functions above to accomodate non-constant payments:
```{r}
P_i = Vectorize(
  # Compute outstanding principal after i payments
  function(A_seq, P0, r, i) {
    # A_seq -- sequence of payments (length m)
    if (i==0) return(P0)
    (1 + r)^i * P0 - sum(A_seq[1:i] * (1 + r)^(0:(i-1)))
  },
  vectorize.args = 'i')

gen_A_seq = function(P0, r, m, Term=36) {
  # Sequence of m - 1 installments, plus final payment (1+r)*P_{m-1}
  Inst = A(P0, r, Term)
  A_seq = rep(Inst, m - 1)
  c(A_seq, (1 + r) * P_i(A_seq, P0, r, m - 1))
}

R1 = Vectorize(
  function(s, P0, r, m) {
    # s -- service fee rate (monthly, on remaining principal)
    # P0, r, m -- args to A()
    A_seq = gen_A_seq(P0, r, m)
    ROI = (sum(A_seq) - s * sum(P_i(A_seq, P0, r, 1:m)) - P0) / P0
    # annualize
    (1 + ROI)^(12/m) - 1
  },
  vectorize.args = 'm')

R2 = Vectorize(
  function(s, P0, r, m=36) {
    # s  -- service fee rate (monthly, on payments)
    # P0, r, m -- args to A()
    A_seq = gen_A_seq(P0, r, m)
    ROI = ((1 - s) * sum(A_seq) - P0) / P0
    # annualize
    (1 + ROI)^(12/m) - 1
  },
  vectorize.args = 'm')

# Returns for institutional (whole loan) investors
r_inst = R1(s=.013/12, P0=1000, r=.15/12, m=1:36)

# Returns for retail (Note) investors
r_ret  = R2(s=.01, P0=1000, r=.15/12, m=1:36)

matplot(1:36, cbind(r_inst, r_ret), las=1, type='l',
        xlab='Months until full payment (m)',
        ylab='Annualized returns')
legend('topright', c('R1','R2'), lty=c(1,2), col=c('black','red'),
       inset=.01)
```

With this more realistic payment sequence, an investor would still prefer early payment under $R_1$, though the "penalty" on returns of one additional month---corresponding to the slope of the curve---is not as pronounced for small $m$. Interestingly, the curve corresponding to $R_2$ attains an interior maximum, suggesting that a Note investor might be best off if the borrower prepays, but not _too_ early.

