---
title: "Tips and Tricks with Reduce"
date: 2017-08-31
categories: ["R"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
```

```{r echo=FALSE}
set.seed(30)
```

R comes with a number of built-in functional programming tools, one of which is the `Reduce` function which recursively calls a binary function over a vector of inputs. It's quite versatile, but most explanations of `Reduce` tend to focus on trivial use cases that are not particularly inspiring. Even Hadley's [_Advanced R_](http://adv-r.had.co.nz/) (a fantastic reference!) limits the discussion to implementations of cumulative sum---for which there is already `cumsum`---and multiple intersect. I'll use this post to demonstrate some less obvious examples of how `Reduce` can come in handy, based on an alternative interpretation of the function as a simple kind of [state machine](https://ocw.mit.edu/courses/electrical-engineering-and-computer-science/6-01sc-introduction-to-electrical-engineering-and-computer-science-i-spring-2011/unit-1-software-engineering/state-machines/MIT6_01SCS11_chap04.pdf). The examples aren't necessarily applicable to day-to-day data analysis tasks, but should be interesting enough to impart a more expansive understanding of what `Reduce` can do.

If you've never heard of `Reduce` or have never thought to use it before, I recommend taking a quick look at [this](http://adv-r.had.co.nz/Functionals.html#functionals-not) or [this](http://www.brodrigues.co/fput/fprog.html#reduce) or [this](https://stackoverflow.com/questions/28545688/understand-the-reduce-function) before reading further.

I think it's appropriate to start by revisiting an example from _Advanced R_ where Hadley claims it is _not_ possible to use a functional. Of a for-loop implementation of an [exponential moving average](https://en.wikipedia.org/wiki/Moving_average#Exponential_moving_average) (also _exponential smooth_), he writes, "We can't eliminate the for-loop because none of the functionals we've seen allow the output at position `i` to depend on both the input and output at position `i-1`" ([Chapter 11, page 226](http://adv-r.had.co.nz/Functionals.html#functionals-not)).

This is precisely what `Reduce` does! In fact, given `x` and smoothing parameter `alpha`, we can compute the EMA without using an explicit for-loop^[Of course, if you take a look at the source code for `Reduce` you'll see that it uses a for-loop under the hood.]:

```{r}
exps <- function(x, alpha) {
  Reduce(function(s, x_) alpha*x_ + (1-alpha)*s, x, accumulate=TRUE)
}

x <- cumsum(rnorm(100))

plot(x, type='l')
lines(exps(x, alpha=.2), col='red')
lines(exps(x, alpha=.4), col='blue')
```

(As a reminder, specifying `accumulate=TRUE` causes `Reduce` to return all intermediate results, so that the output has the same length as `x`. We can also specify an initial value with the `init` argument, in which case the output will have length `length(x) + 1`.)


More generally, `Reduce` can implement a state process where the state is updated according to its previous value and an input (or an "action"), i.e., `$ s' = f(s, a) $`.^[The process is restricted to be _Markov_, i.e., dependent only on the input `a` and the "current" `s`, and not on any previously computed values of `s`.] In the EMA example, the state is `s` and the input is `x_`, an element of `x`. This conception of `Reduce` is related to the idea of applying a binary function recursively, but perhaps provides a more natural way of thinking about certain kinds of problems.

To see this in a second example, suppose that we'd like to compute a cumulative sum that "resets" after eating the value zero. All we need to do is write a function that takes the current sum (the state) as given, and either adds to or resets that sum depending on the value of the input:
```{r}
x <- c(1, 2, 4, 0, -1, 3, 0, 1, 1, 1)

Reduce(function(s, x_) (s + x_)*(x_ != 0), x, accumulate=TRUE)
```
Easy!

We can also use `Reduce` to implement a version of `zoo::na.locf` ("last observation carry forward"), which replaces `NA` with the last non-`NA` value:

```{r}
x <- rep(NA_integer_, 14)
x[sample(length(x), 5)] <- 1:5
x

Reduce(function(s, x_) if (is.na(x_)) s else x_, x, accumulate=TRUE)
```

The function simply returns `s` if the input `x_` is `NA`; otherwise `x_` becomes the new `s`.

Now suppose we have a vector of 0, 1, and -1 which represent actions peformed on some state which can be 0 or 1. If the action "1" is performed, the state switches to (or remains) "1"; if "0" is performed, there is no effect; and if "-1" is performed, the state switches to (or remains) "0". (Think of this as flipping a light switch on or off.) We can use `Reduce` to compute the values of the state given a sequence of actions `A` and an initial state:

```{r}
(A <- sample(c(-1,0,1), size=14, replace=TRUE))

Reduce(function(s, a_) as.integer(s + a_ > 0), A, init=1L, accumulate=TRUE)
```

I've used this as a method to strip away blocks^[I.e., multiple lines in a data frame after using `readLines`] of unnecessary text (headers, footers, etc.) from multi-page PDFs after text conversion. All I need to do is mark the rows where the useful data and the useless text begin, then `Reduce` and drop.

All of the examples so far have used integer states and inputs, but it's straightforward to work with more complex object types like matrices and data frames. We could use `Reduce` to merge a list of data frames (a common data analysis task), or to simulate sample paths for a multivariate time series where the state is a vector `$S$`.

There's even a way to make `Reduce` handle multiple input arguments. Say we want to recursively compute `$ s' = a_1 \times s + a_2 $` given a sequence of `$ (a_1, a_2) $` pairs. We can use `Map` (another useful functional) to combine `A1` and `A2` pairwise and then use standard indexing to access the different inputs:
```{r}
A1 <- rnorm(10)
A2 <- rnorm(10)

Reduce(function(s, a) a[1] * s + a[2], Map(c, A1, A2), init=1L, accumulate=TRUE)

# or:
# Reduce(function(s, a) rnorm(1) * s + rnorm(1), seq_len(10), init=1L, accumulate=TRUE)
```

One last example from [StackOverflow](https://stackoverflow.com/questions/41073261/). Say we have a binary vector `x` and want to replace the two elements following every "1" with "1". We can use `Reduce` to create a counter that decrements by 1 in the case of "0" and resets to 3 in the case of "1", so that any two "0" elements following a "1" correspond to positive values in the counter. Taking the `sign` then gives the desired result:
```{r}
x <- rep(0, 14)
x[sample(length(x), 5)] <- 1
x

(counter <- Reduce(function(s, x_) min(3, max(s-1, 0) + x_), 3*x, accumulate=TRUE))

sign(counter)
```

This solution takes some thought, and perhaps an alternative method would be preferable for the sake of readability. But the point is that `Reduce` is flexible enough to be a contender for a much wider range of problems than is usually recognized. My hope is that the above examples help illuminate how `Reduce` (and functional programming) can be used cleverly to provide compact solutions for certain kinds of programming problems.
